---
title: "Introduction"
author: "박선철 (SNU)"
date: "November 8, 2015"
header-includes:
  - \usepackage{bm}
  - \usepackage{kotex}
output: html_document
---

## About the course

- Plan:

- Useful references

1. Coles (2001), *An introduction to Statistical Modeling of Extrem Values*, Springer

2. Beirlant, Goegebeur, Segers and Teugels (2004), *Statistics of Extremes: Theory and Applications*, Wiley

3. de Haan and Ferreira (2006), *Extreme Value Theory: An Introduction*, Springer

4. Dey and Yan (2016), *Extreme Value Modeling and Risk Analysis*, CRC Press

- `R` packages: `evd`, `ismev`, `POT`, `RandomFields`, `SpatialExtremes`, etc

## Intro

## Examples: Daily rainfal accumulations

```{r, echo=F, fig.cap='A time series of daily rainfall accumulations at a location in south-west England recorded over 1914-1962.', message=F, warning=F, cache=F, out.width = "275px", dev.args=list(pointsize=20)}
library(ismev)

par(mfrow=c(1,1))
data("rain")
plot(seq(ISOdate(1914,1,1), ISOdate(1961,12,30), "days"), rain, xlab="Year", ylab="Daily Rainfall")

```


## Examples: Wooster data

```{r, echo=F, fig.cap='Negated Wooster 5-year daily minimum temperatures.', message=F, warning=F, cache=F, out.width = "275px", dev.args=list(pointsize=20)}

par(mfrow=c(1,1))
data("wooster")
plot(seq(ISOdate(1983,1,1), ISOdate(1987,12,31), "days"), -wooster, xlab="Year", ylab="Daily Minimum Temp.")

```


## Challenges

- Modeling of the **sizes** of extreme events at some specific location (pointwise return levels) — Block maximum/Threshold approaches

- Modeling of **clusters** of extremes (due to **temporal dependence**)

- **Joint** modeling of extremes at several locations (**spatial dependence**)

- Modeling of **non-stationarity**

- Models for **asymptotic independence**

- **Measure of dependence** at extreme levels

- **Efficient inference**

- **Simulation** of extreme events in space and time
